Understanding the Impact of Mortgage State and Income on Household's Poverty Status

### 2024-03-30 paper/paper.qmd
finish model and result

---
title: "Understanding the Impact of Marital Status, Income, and Age on Individuals' Poverty Status"
subtitle: "An Analysis of the 2019 Revised Supplemental Poverty Measure Dataset"
author: 
  - Mary Cheng
thanks: "Code and data are available at: https://github.com/marycx/us_poverty_analysis_2019.git"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

@@

@@

## Data Source

## Features

@@

@@

poverty2019 <-
  poverty2019 |>
  mutate(
    poverty_status = as_factor(poverty_status),
    mortgage_state= factor(
      mortgage_state,
      levels = c(
      "Owner with Mortgage",
      "Owner without Mortgage or rent-free",
      "Renter")
    ),
    income = factor(
      income,
      levels = c(
      "below 10k",
      "10k-50k",
      "50k-100k",
      "100k-150k",
      "150k-200k",
      "200k-250k",
      "above 250k")
    )
  ) |>
  select(poverty_status, mortgage_state, income)
```

@@

poverty2019[5:9, 1:4] |>
  kable()
```

@@

poverty2019 |>
  summary() |>
  kable(align = "c") %>%
  kable_styling(font_size = 8)
```

## Data Visualization
```{r}
#| label: fig-mortgage
#| fig-cap: The distribution of poverty status by mortgage state
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = mortgage_state, fill = poverty_status)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  theme_minimal() +
  theme(axis.text.x=element_text(hjust=1, vjust=0.5),
        text=element_text(size=6)) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Mortgage state",
    y = "Number of respondents",
    fill = "Poverty status"
  )
```
```{r}
#| label: fig-income
#| fig-cap: The distribution of poverty status by income level
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = income, fill = poverty_status)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Income level",
    y = "Number of respondents",
    fill = "Poverty status"
  )
```

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\pi_i &\sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \alpha + \beta_1 \times \mbox{mortgage}_i + \beta_2 \times \mbox{income}_i + \beta_3 \times \mbox{age}_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

@@

## First discussion point {#sec-first-point}

@@

## Second discussion point

## Third discussion point

@@

@fig-trace1, @fig-trace2, @fig-trace3 are the trace plots of the model. It tells us if there is existence of signs that the our model runs into issues. We observe lines in all the trace plots are horizontal and oscillating, and have overlaps between the chains. This suggests that there is nothing strange in this trace plot.

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace1
#| fig-cap: "Trace plot of intercept and marital status"
#| fig-subcap: ["Trace plot of Intercept", "Trace plot of Owner without Mortgage or rent-free", "Trace plot of Renter"]
#| layout-ncol: 2

plot(poverty_prediction_model, "trace", "(Intercept)")
plot(poverty_prediction_model, "trace", "mortgage_stateOwner without Mortgage or rent-free")
plot(poverty_prediction_model, "trace", "mortgage_stateRenter")
```

@@

plot(poverty_prediction_model, "rhat")
```
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-modelresults1
#| fig-cap: "Credible intervals for predictors of positive poverty status"

modelplot(poverty_prediction_model, conf_level = 0.90, size = 0.2) +
  labs(x = "90% credibility interval")
```

@@

# Modify the x-axis limits
model_plot + xlim(-5, 5) +  # Adjust the limits as needed
  labs(x = "90% Credibility Interval")
```

>>>>

---
title: "Understanding the Impact of Mortgage State and Income on Household's Poverty Status"
subtitle: "An Analysis of the 2019 Revised Supplemental Poverty Measure Dataset"
author: 
  - Mary Cheng
thanks: "Code and data are available at: https://github.com/marycx/us_poverty_analysis_2019.git"
date: today
date-format: long
abstract: "This paper examines household poverty status in the United States in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research Dataset from the United States Census Bureau. Using logistic regression analysis, we investigate the influence of household income level and mortgage state on poverty status. Our findings reveal a higher likelihood of poverty for households earning less than $50000 and for renters compared to property owners. Conversely, all households with incomes exceeding $100000 are not in poverty. These insights emphasize the need for targeted interventions to address socioeconomic inequalities and promote financial well-being among US households."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

\newpage
# Introduction

@@

## Estimand

@@

## Data Source
This report uses the 2019 Revised Supplemental Poverty Measure (SPM) Research Dataset provided by the United States Census Bureau as our main source of data. The Census Bureau provides information about the people and economy of the United States, with a goal to support economic growth, enhance scientific knowledge, and assist in making informed decisions. The Supplemental Poverty Measure (SPM), similar to the official poverty measure, tackles the shortcomings of poverty assessment. It determines poverty status by gauging resources against a defined standard of living. This standard considers expenditures on essentials like food, clothing, shelter, and utilities, plus a bit more for additional expenses. The resources assessed include disposable income, accounting for taxes and some noncash benefits, available to cover these needs. In 2019, the dataset contained a total of 157,959 entries.

## Features
The original SPM 2019 dataset, which shows in @tbl-raw, contains 157959 data entries and many variables. Since it is difficult to observe such a large dataset, this report will only explore and analyze through several data features. We chose these 4 variables: "h_seq", "spm_poor", "spm_tenmortstatus", "spm_totval", in our analysis.

1. h_seq: Household sequence number, unique identifier for each household.

2. spm_poor: the poverty status; 1 representing "In poverty" and 0 represneting "Not in poverty".

3. spm_tenmortstatus: household's tenure/mortgage status; 1 representing "Owner with Mortgage", 2 representing "Owner without Mortgage or rent-free", and 3 representing "Renter".

4. spm_totval: household's cash income

@@

The original dataset includes duplicate responses from the same household. As the dataset features are related to the entire household as a whole, this paper will eliminate duplicate entries, using the variable "h_seq" which is an unique household identifier, to ensure data accuracy and consistency.

The dataset is cleaned by renaming of column headers, defining column classes, and replacing numerical values in the table with their corresponding descriptions from the data dictionary to improve the readability. After cleaning, 62917 rows of data with 3 data features remain. @tbl-clean-data shows a preview of the cleaned dataset.

@@

poverty2019 <-
  poverty2019 |>
  mutate(
    poverty_status = as_factor(poverty_status),
    mortgage_state= factor(
      mortgage_state,
      levels = c(
      "Owner with Mortgage",
      "Owner without Mortgage",
      "Renter")
    ),
    income = factor(
      income,
      levels = c(
      "below 10k",
      "10k-50k",
      "50k-100k",
      "100k-150k",
      "150k-200k",
      "200k-250k",
      "above 250k")
    )
  ) |>
  select(poverty_status, mortgage_state, income)
```

@@

poverty2019[1:5, 1:3] |>
  kable()
```

@tbl-summary is a summary of the cleaned data, showing detailed statistics about the dataset. As we can see from the table, there are more households which are not in poverty. Respondents cover a wide range of income levels, with "50k-100k" and "100k-150k" being more heavily represented. Also, the mortgage states of the households vary, each of the three mortgage states being fairly evenly represented.

@@

poverty2019 |>
  summary() |>
  kable(align = "c") %>%
  kable_styling(font_size = 12)
```

## Data Visualization
@fig-mortgage illustrates the relationship between mortgage state and poverty status of households. It can be seen that renters constitute the highest proportion of households experiencing poverty, followed by owners without mortgages, and finally owners with mortgages. Conversely, for households who are not in poverty, owners with mortgages represent the largest segment, followed by owners without mortgages, and then renters. In addition, it can be observed that overall a larger proportion of households are not experiencing poverty. Despite variations in poverty rates across different mortgage states, the cumulative data shows economic stability among the majority of households in the United States in 2019.

```{r}
#| label: fig-mortgage
#| fig-cap: The distribution of poverty status by mortgage state
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = poverty_status, fill = mortgage_state)) +
  geom_bar(width = 0.2) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Poverty Status",
    y = "Number of respondents",
    fill = "Mortgage state"
  )
```

@fig-income shows the relationship between household annual total income and their poverty status. We can see that among households earning more than 100k annually, no household is in poverty. For household with income between 50k to 100k, almost all of them are not in poverty with only a very small number of households in poverty. Within the income range of 10k-50k, the majority of households are not in poverty, but there are some families in poverty. However, for households earning below 10k annually, almost all of them are in poverty. This indicates a strong positive correlation between higher income levels and financial stability.

```{r}
#| label: fig-income
#| fig-cap: The distribution of poverty status by income level
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = income, fill = poverty_status)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Income level",
    y = "Number of respondents",
    fill = "Poverty status"
  )
```

\newpage
# Model
In our analysis, we utilized a Bayesian logistic regression model to examine the relationship between poverty status and two demographic factors, mortgage status and income of the household. Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up
The model is formulated as follows:
\begin{align} 
y_i|\pi_i &\sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \alpha + \beta_1 \times \mbox{mortgage}_i + \beta_2 \times \mbox{income}_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5)
\end{align}

In this model, $y_i$ represents the binary outcome variable indicating whether a household is in poverty (as opposed to not in poverty). The probability of being in poverty ($\pi_i$) is modeled using a logistic link function (\text{logit}($\pi_i$)), which is a linear combination of the intercept ($\alpha$) and the coefficients ($\beta_1$, $\beta_2$) corresponding to the predictor variables, mortgage state and income level, respectively. These predictor variables are denoted as income_i and mortgage_state_i, where $i$ indexes the individual households in the dataset.

The intercept ($\alpha$) and coefficients ($\beta_1$, $\beta_2$) are assigned informative prior distributions to regularize the model. Specifically, we assume a normal distribution with a mean of 0 and a standard deviation of 2.5 for each parameter.

We chose this modeling approach for several reasons. Firstly, logistic regression is ideal for binary outcome variables, making it appropriate for analyzing poverty status. Also, Bayesian methods enable us to integrate prior knowledge and uncertainty into the analysis, resulting in more reliable estimates of the model parameters.

While alternative modeling approaches like linear regression were considered, we selected Bayesian logistic regression due to the binary nature of our outcome variable, poverty status.

We use the `rstanarm` package [@citeRstanarm] in R [@citeR] to run the model. Default priors from `rstanarm` is used. Rstanarm employs Markov chain Monte Carlo (MCMC) techniques to estimate the posterior distribution of the parameters. To avoid excessive runtime, 1000 data entries are randomly sampled to fit the model with random seed 215. Model diagnostics, including convergence checks and posterior summaries, are available in the supplementary materials (see Appendix @sec-model-details).

### Model justification
Regarding the relationship between mortgage state and poverty state, we anticipate that both owners without mortgages and owners with mortgages are less likely to experience poverty. Owners without mortgages typically indicate financial stability, as they have already paid off their homes and can allocate financial resources toward household necessities, other than paying back the mortgage / rent. Also, without monthly mortgage or rent payments, their financial burden is reduced, further decreasing the likelihood of poverty. Similarly, owners with mortgages demonstrate financial responsibility by securing loans from banks. The approval of these loans suggests that the bank is confident in owner's ability to manage repayments, indicating a lower possibility of poverty. On the other hand, renters are more likely to experience poverty due to their lack of property ownership and the hidden implication that they do not earn enough to purchase their own property. Renters typically have fewer savings and may earn insufficient salaries to cover rental and essential expenses. Consequently, the combination of rental payments and other financial obligations increases their vulnerability to poverty.

In terms of income levels, we expect a positive relationship between household's income level and their poverty status. This expectation arises from common sense that higher income typically reduces the likelihood of experiencing poverty. If a family is earning a lot of money, then naturally it would not be in poverty. Households that are earning less than 10k annually in the United States obviously have a larger chance in being in poverty due to the high cost of living. Their income probably barely covers or cannot cover for essential expenses such as housing, transportation, or food.

## Model Implication
For posterior predictive checks, in @fig-post_dist, the great fit of the posterior distribution from our logistic regression model with actual poverty data suggests accurate capture of poverty status patterns. This indicates the accuracy of our model's poverty status prediction to the 2019 SPM data. Also, @fig-post_prior compares the posterior to the prior, which shows some parameter changes such as "below 10k" and "10k-50k", as well as the intercept (people who are owners with mortgage and earn 100-150k). The discrepancy for the parameter suggest that our prior may not be very accurate in regards to these specific aspects.

The trace plots in @fig-trace1 and @fig-trace2 do not suggest anything out of the ordinary. Also, with the Rhat plot (@fig-rhat), we can observe that everything is close to 1, and no more than 1.05, which shows the great convergence of Markov chain Monte Carlo for our model.

More detailed explanation of each plot can be found in Appendix @sec-model-details.

# Results
Our results are summarized in @tbl-modelresults. Our results generally matches our expectation. To avoid multicollinearity, the model excludes one variable from each category: mortgage state "Owner with mortgage" and income level "100k-150k". The intercept represents the estimated log-odds of being in poverty when all other predictors are held constant at their reference levels. In this case, the estimated log-odds of being in poverty for people who are owner with mortgage and their total household income level to be 100k-150k annually is −6.782.

The possibility of households with income level below 10k being in poverty is large. The estimated coefficient of 17.653 suggests that, holding all other variables constant, households with income less than 10k are estimated to have a 17.653 unit increase in the log-odds of being in poverty compared to the reference group. Households with income level at "10k-50k" and "50k-100k" on average are also more likely to be in poverty compared to the reference group, with the estimated coefficient to be 5.478 and 1.640 respectively.

The mortgage status of individuals also influences their poverty status. As expected, renters exhibit a higher likelihood of experiencing poverty, as shown by the estimated coefficient of 0.359. However, it's important to note that while the coefficient is positive compared to the reference group, its magnitude is relatively small, suggesting a moderate rather than a substantial difference.

The mortgage status of individuals also influences their poverty status. As expected, renters exhibit a higher likelihood of experiencing poverty, as shown by the estimated coefficient of 0.359. However, it's important to note that while the coefficient is positive compared to the reference group, its magnitude is relatively small, suggesting a moderate rather than a substantial difference.

@fig-modelresults1 (see @sec-credibility-interval) shows range of coefficient estimates of our model within the 90% probability. Due to the fact that the credibility interval for mortgage "Renter" and "Owners without mortgage" is quite small, it is hard to observe the trend of the 90% credibility intervals of these two variables. Therefore we created @fig-modelresults2 with the x axis limited from -5 to 5.

Combining @fig-modelresults1 and @fig-modelresults2, we observe statistical significance for the coefficient estimates for household with income below 10k, household income between 10k to 50k, and the intercept, household which owns a property with mortgage and has a income level between 100k to 150k. The estimates are significant because their credibility intervals do not cross 0. The value for the estimates are in log-odds, indicating that if the coefficient is positive, the household is in poverty, if negative, the household is not in poverty.

@@

## Relationship betwen Mortgage State and Poverty Status

@@

## Relationship betwen Income and Poverty Status

Average annual expenditure $63,036

## mortgage interest rate etc...

@@

@fig-trace1 and @fig-trace2 are the trace plots of the model. It tells us if there is existence of signs that the our model runs into issues. We observe lines in all the trace plots are horizontal and oscillating, and have overlaps between the chains. This suggests that there is nothing strange in this trace plot.

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace1
#| fig-cap: "Trace plot of intercept and marital status"
#| fig-subcap: ["Trace plot of Intercept", "Trace plot of Owner without Mortgage", "Trace plot of Renter"]
#| layout-ncol: 2

plot(poverty_prediction_model, "trace", "(Intercept)")
plot(poverty_prediction_model, "trace", "mortgage_stateOwner without Mortgage")
plot(poverty_prediction_model, "trace", "mortgage_stateRenter")
```

@@

plot(poverty_prediction_model, "rhat")
```

## 90% Credibility Interval {#sec-credibility-interval}
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-modelresults1
#| fig-cap: "Credible intervals for predictors of positive poverty status"

modelplot(poverty_prediction_model, conf_level = 0.90, size = 0.2) +
  labs(x = "90% credibility interval") +
  theme(axis.text.y = element_text(size = 6))
```

@@

# Modify the x-axis limits
model_plot + xlim(-5, 5) +  # Adjust the limits as needed
  labs(x = "90% Credibility Interval") +
  theme(axis.text.y = element_text(size = 6))
```
